---
title: "Plague"
format: html
description: test description
execute:
  eval: true
  warning: false
---

Single cell RNA-seq of PBMCs were collected at baseline and stimulated with plaugue. 
The effect of stimulation was estimated in each celltype using pseudobulk expression. 
The resutling effect sizes across cell types were restimated using MASH. Differentially expressed genes in a given cell type are identified as those with $\text{lfsr} < 0.01$.

```{r load-data}
library(dplyr)
library(AnnotationDbi)
library(org.Hs.eg.db)

reticulate::use_virtualenv('gibss')

data <- new.env()
load('data/plague.RData', env=data)
```


One thing to note is that the gene lists are quite large when using an lfsr threshold of 0.01, and the DE genes in each cell type have substantial overlap.

```{r intersection-fun}
make_intersection_matrix <- function(list_of_lists){
  # Function to calculate the size of the intersection of two lists
  intersection_size <- function(a, b) {
    length(intersect(a, b))
  }
  
  # Determine the number of lists
  n <- length(list_of_lists)
  
  # Create an empty matrix to store the intersection sizes
  intersection_matrix <- matrix(0, n, n, dimnames = list(names(list_of_lists), names(list_of_lists)))
  
  # Calculate the size of the intersection for each pair
  for (i in seq_along(list_of_lists)) {
    for (j in seq_along(list_of_lists)) {
      # Fill in the matrix with intersection sizes, ensuring i <= j to avoid redundant calculations
      if (i <= j) {
        intersection_matrix[i, j] <- intersection_size(list_of_lists[[i]], list_of_lists[[j]])
      }
    }
  }
  
  # Since the intersection operation is symmetric, mirror the upper triangle to the lower triangle
  intersection_matrix[lower.tri(intersection_matrix)] <- t(intersection_matrix)[lower.tri(intersection_matrix)]
  return(intersection_matrix)
}
```


```{r cs-tbl-funs}
make_cs_tbl <- gseasusie:::make_cs_tbl
make_component_tbl <- gseasusie:::make_component_tbl

make_cs_tbl_nested <- function(fit, logbf_threshold=log(10.)) {
  more_go_info <- AnnotationDbi::select(GO.db::GO.db,
    keys = unique(fit$data$geneSets$geneSet),
    columns = c("TERM", "DEFINITION"),
    keytype = "GOID"
  ) %>%
    as_tibble() %>%
    dplyr::mutate(geneSet = GOID)

  more_gene_info <- AnnotationDbi::select(org.Hs.eg.db::org.Hs.eg.db,
    keys = unique(fit$data$geneMapping$gene),
    columns = c("SYMBOL", "GENETYPE", "GENENAME"),
    keytype = "ENTREZID"
  ) %>%
    as_tibble() %>%
    dplyr::mutate(gene = ENTREZID)

  gene_columns <- c("ENTREZID", "SYMBOL", "GENENAME", "geneInList")
  gene_set_columns <- c("geneSet", "TERM", "DEFINITION", "alpha", "beta", "lbf", "geneSetSize", "propInList")
  component_columns <- c("component", "size", "top_term", "top_effect", "lbf_ser")

  all_columns <- c(component_columns, gene_set_columns, gene_columns)

  message("building nested credible set table")
  cs_tbl <- make_cs_tbl(fit, logbf_threshold)
  component_tbl <- make_component_tbl(fit, logbf_threshold)
  nested_tbl <- NULL
  if(nrow(cs_tbl) > 0){
    nested_tbl <- cs_tbl %>%
      left_join(component_tbl) %>% # add effect estimates, etc.
      left_join(more_go_info) %>%
      left_join(more_gene_info) %>%
      dplyr::select(tidyselect::any_of(all_columns)) %>%
      # get representative gene set info
      dplyr::group_by(component) %>%
      mutate(
        top_term = TERM[which.max(lbf)],
        top_effect = beta[which.max(lbf)]
      ) %>%
      ungroup() %>%
      # nest gene level data
      tidyr::nest(.by = c(component_columns, gene_set_columns), .key = "details") %>%
      # nest gene set level data
      tidyr::nest(.by = component_columns, .key = "details")
  }
  return(nested_tbl)
}
```


```{r reactable-funs}
library(reactable)

make_sensible_colDef <- function(data) {
  # Initialize an empty list to store colDef objects
  col_defs <- list()

  # Iterate over each column in the data
  for (col_name in names(data)) {
    # Get the column data
    col_data <- data[[col_name]]
    
    # Determine the type of the column
    col_type <- class(col_data)
    
    # Set default colDef based on the column type
    if ("numeric" %in% col_type) {
      # Numeric columns: Format to 2 decimal places
      col_defs[[col_name]] <- colDef(format = colFormat(digits = 2))
    } else if ("integer" %in% col_type) {
      # Integer columns: No decimal places
      col_defs[[col_name]] <- colDef(format = colFormat(digits = 0))
    } else if ("character" %in% col_type || "factor" %in% col_type) {
      # Character or factor columns: Truncate long strings
      # col_defs[[col_name]] <- colDef(truncate = TRUE, minWidth = 150)
    } else if ("Date" %in% col_type) {
      # Date columns: Format as a readable date
      col_defs[[col_name]] <- colDef(format = colFormat(date = TRUE))
    } else if ("logical" %in% col_type) {
      # Logical columns: Display as checkboxes
      col_defs[[col_name]] <- colDef(cell = function(value) {
        if (isTRUE(value)) "✔" else "✗"
      })
    } else {
      # Default colDef for other types
      col_defs[[col_name]] <- colDef()
    }
  }
  return(col_defs)
}

make_rwg_column_style <- function(max=2){
  function(values){
        pal <- function(x) rgb(colorRamp(c('red', 'white', 'green'))(x), maxColorValue=255)
        clipped <- (pmax(-max, pmin(values, max)) + max) / (2*max)
        color <- pal(clipped)
        list(background = color)
  }
}

make_column_format_list <- function(data){
  columns <- make_sensible_colDef(data)
  if ('top_effect' %in% colnames(data)){
    columns$top_effect <- reactable::colDef(
      style= make_rwg_column_style(2),
      format = reactable::colFormat(
        digits=2
      )
    )
  }
  if ('beta' %in% colnames(data)){
    columns$beta <- reactable::colDef(
      style= make_rwg_column_style(2),
      format = reactable::colFormat(
        digits=2
      )
    )
  }
  if ('alpha' %in% colnames(data)){
    columns$alpha <- reactable::colDef(
      format = reactable::colFormat(digits=2)
    )
  }
  return(columns)
}

make_get_details <- function (nested_tbls) 
{
    if (!is.null(nested_tbls)) {
      get_details <- function(index) {
        nested_tbl <- nested_tbls[[index]]
        data <- dplyr::select(nested_tbl, -tidyselect::any_of("details")) %>% head(10)
        if('details' %in% colnames(nested_tbl)){
          details <- nested_tbl$details
        } else{
          details <- NULL
        }
        htmltools::div(style = "padding: 1rem", reactable::reactable(
          data,
          details = make_get_details(details), 
          outlined = TRUE, 
          columns = make_column_format_list(data)
        ))
      }
    }
    else {
        get_details <- NULL
    }
    return(get_details)
}

nested_reactable <- function (nested_tbl) 
{
  if(is.null(nested_tbl)){
    return(NULL)
  }
  data <- dplyr::select(nested_tbl, -tidyselect::any_of("details"))
  details <- nested_tbl$details
  reactable::reactable(
    data, 
    details = make_get_details(details),
    columns = make_column_format_list(data)
  )
}
```

## DE genes (`lfsr < 0.01`)

### Overview

```{r}

make_list_and_background <- function(gene, lfsr){
  gene_list <- gene[which(lfsr < 0.01)]
  return(list(gene_list=gene_list, gene_background=gene))
}

symbol2entrez <- AnnotationDbi::select(
  org.Hs.eg.db, 
  keys = unique(data$gene_results$gene), 
  columns = 'ENTREZID', 
  keytype = 'SYMBOL') %>%
  dplyr::rename(gene = SYMBOL)

gene_lists <- data$long_results %>%
  dplyr::left_join(symbol2entrez) %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::group_by(celltype) %>%
  dplyr::summarise(gene_list = list(make_list_and_background(ENTREZID, lfsr)))
```

```{r heatmap-intersection}
# Display the matrix
list_of_lists <- purrr::map(gene_lists$gene_list, ~.$gene_list)
names(list_of_lists) <- gene_lists$celltype
intersection_matrix <- make_intersection_matrix(list_of_lists)
pheatmap::pheatmap(intersection_matrix, cluster_rows = FALSE, cluster_cols = FALSE, display_numbers = TRUE)
```

### Fitting

```{r fit1}
#| cache: true

susie_fits <- gene_lists %>%
  dplyr::rowwise() %>%
  mutate(fit = list(gseasusie::fit_gsea_susie_webgestalt(
    gene_list$gene_list, gene_list$gene_background,
    'geneontology_Biological_Process'
  )))
```

### B Cell

```{r b}
susie_fits$fit[[1]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### CD4T Cell

```{r cd4t}
susie_fits$fit[[2]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### CD8T Cell

```{r cd8t}
susie_fits$fit[[3]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### Mono

```{r mono}
susie_fits$fit[[4]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### NK

```{r nk}
#| eval: True 
susie_fits$fit[[5]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```


## DE genes (`lfsr < 0.01`, `log2FC > 0.5`)

### Overview

```{r prepare-lists-large-effect}
make_list_and_background <- function(gene, lfsr, beta){
  gene_list <- gene[intersect(which(lfsr < 0.01), which(abs(beta *log2(exp(1))) > 0.5))]
  return(list(gene_list=gene_list, gene_background=gene))
}

symbol2entrez <- AnnotationDbi::select(
  org.Hs.eg.db, 
  keys = unique(data$gene_results$gene), 
  columns = 'ENTREZID', 
  keytype = 'SYMBOL') %>%
  dplyr::rename(gene = SYMBOL)

gene_lists <- data$long_results %>%
  dplyr::left_join(symbol2entrez) %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::group_by(celltype) %>%
  dplyr::summarise(gene_list = list(make_list_and_background(ENTREZID, lfsr, beta)))
```


```{r heatmap-intersection-large-effect}
# Display the matrix
list_of_lists <- purrr::map(gene_lists$gene_list, ~.$gene_list)
names(list_of_lists) <- gene_lists$celltype
intersection_matrix <- make_intersection_matrix(list_of_lists)
pheatmap::pheatmap(intersection_matrix, cluster_rows = FALSE, cluster_cols = FALSE, display_numbers = TRUE)
```

### Fitting

```{r fit-large-effect}
#| cache: true

susie_fits <- gene_lists %>%
  dplyr::rowwise() %>%
  mutate(fit = list(gseasusie::fit_gsea_susie_webgestalt(
    gene_list$gene_list, gene_list$gene_background,
    'geneontology_Biological_Process'
  )))
```

### B Cell

```{r b-large-effect}
susie_fits$fit[[1]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### CD4T Cell

```{r cd4t-large-effect}
susie_fits$fit[[2]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### CD8T Cell

```{r cd8t-large-effect}
susie_fits$fit[[3]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### Mono

```{r mono-large-effect}
susie_fits$fit[[4]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### NK

```{r nk-large-effect}
#| eval: True 
susie_fits$fit[[5]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```



## DE genes (`lfsr < 0.01`, `log2FC > 0.2`, upregulated)

### Overview

```{r prepare-lists-large-effect-up}
make_list_and_background <- function(gene, lfsr, beta){
  gene_list <- gene[intersect(which(lfsr < 0.01), which(beta *log2(exp(1)) > 0.5))]
  return(list(gene_list=gene_list, gene_background=gene))
}

symbol2entrez <- AnnotationDbi::select(
  org.Hs.eg.db, 
  keys = unique(data$gene_results$gene), 
  columns = 'ENTREZID', 
  keytype = 'SYMBOL') %>%
  dplyr::rename(gene = SYMBOL)

gene_lists <- data$long_results %>%
  dplyr::left_join(symbol2entrez) %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::group_by(celltype) %>%
  dplyr::summarise(gene_list = list(make_list_and_background(ENTREZID, lfsr, beta)))
```


```{r heatmap-intersection-large-effect-up}
# Display the matrix
list_of_lists <- purrr::map(gene_lists$gene_list, ~.$gene_list)
names(list_of_lists) <- gene_lists$celltype
intersection_matrix <- make_intersection_matrix(list_of_lists)
pheatmap::pheatmap(intersection_matrix, cluster_rows = FALSE, cluster_cols = FALSE, display_numbers = TRUE)
```

### Fitting

```{r fit-large-effect-up}
#| cache: true

susie_fits <- gene_lists %>%
  dplyr::rowwise() %>%
  mutate(fit = list(gseasusie::fit_gsea_susie_webgestalt(
    gene_list$gene_list, gene_list$gene_background,
    'geneontology_Biological_Process'
  )))
```

### B Cell

```{r b-large-effect-up}
susie_fits$fit[[1]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### CD4T Cell

```{r cd4t-large-effect-up}
susie_fits$fit[[2]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### CD8T Cell

```{r cd8t-large-effect-up}
susie_fits$fit[[3]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### Mono

```{r mono-large-effect-up}
susie_fits$fit[[4]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### NK

```{r nk-large-effect-up}
#| eval: True 
susie_fits$fit[[5]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

## DE genes (`lfsr < 0.01`, `log2FC > 0.2`, down regulated)

### Overview

```{r prepare-lists-large-effect-down}
make_list_and_background <- function(gene, lfsr, beta){
  gene_list <- gene[intersect(which(lfsr < 0.01), which(beta *log2(exp(1)) < -0.5))]
  return(list(gene_list=gene_list, gene_background=gene))
}

symbol2entrez <- AnnotationDbi::select(
  org.Hs.eg.db, 
  keys = unique(data$gene_results$gene), 
  columns = 'ENTREZID', 
  keytype = 'SYMBOL') %>%
  dplyr::rename(gene = SYMBOL)

gene_lists <- data$long_results %>%
  dplyr::left_join(symbol2entrez) %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::group_by(celltype) %>%
  dplyr::summarise(gene_list = list(make_list_and_background(ENTREZID, lfsr, beta)))
```


```{r heatmap-intersection-large-effect-down}
# Display the matrix
list_of_lists <- purrr::map(gene_lists$gene_list, ~.$gene_list)
names(list_of_lists) <- gene_lists$celltype
intersection_matrix <- make_intersection_matrix(list_of_lists)
pheatmap::pheatmap(intersection_matrix, cluster_rows = FALSE, cluster_cols = FALSE, display_numbers = TRUE)
```

### Fitting

```{r fit-large-effect-down}
#| cache: true

susie_fits <- gene_lists %>%
  dplyr::rowwise() %>%
  mutate(fit = list(gseasusie::fit_gsea_susie_webgestalt(
    gene_list$gene_list, gene_list$gene_background,
    'geneontology_Biological_Process'
  )))
```

### B Cell

```{r b-large-effect-down}
susie_fits$fit[[1]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### CD4T Cell

```{r cd4t-large-effect-down}
susie_fits$fit[[2]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### CD8T Cell

```{r cd8t-large-effect-down}
susie_fits$fit[[3]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### Mono

```{r mono-large-effect-down}
susie_fits$fit[[4]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### NK

```{r nk-large-effect-down}
#| eval: True 
susie_fits$fit[[5]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```




## Unique DE genes (`lfsr < 0.01`)
### Overview

CD4 T-Cells ane mononucleocytes have a large number of unique DE genes, but B, CD8T, and NK cells have very few.
Consequently, logistic SuSiE only reports enrichment in the prior two cell types.

```{r heatmap-intersection2}
# Function to find unique elements in a list compared to others
find_unique_elements <- function(target_list, other_lists) {
  # Combine all other lists into a single vector
  combined_others <- unlist(other_lists)
  # Find elements in the target list that are not in the combined vector of other lists
  setdiff(target_list, combined_others)
}

# Create a new list to store lists of unique elements
unique_elements_list <- lapply(seq_along(gene_lists$gene_list), function(i) {
  # Current target list
  target_list <- list_of_lists[[i]]
  # All other lists except the current one
  other_lists <- list_of_lists[-i]
  # Find unique elements for the current list
  find_unique_elements(target_list, other_lists)
})

# Display the matrix
list_of_lists <- unique_elements_list
names(list_of_lists) <- gene_lists$celltype
intersection_matrix <- make_intersection_matrix(list_of_lists)
pheatmap::pheatmap(intersection_matrix, cluster_rows = FALSE, cluster_cols = FALSE, display_numbers = TRUE)
```

### Fitting
```{r fit2}
#| cache: true

# Assign names to the resulting list of lists
# names(unique_elements_list) <- names(list_of_lists)
unique_susie_fits <- gene_lists %>%
  mutate(unique_list = unique_elements_list) %>%
  dplyr::rowwise() %>%
  mutate(fit = list(gseasusie::fit_gsea_susie_webgestalt(
    unique_list, gene_list$gene_background,
    'geneontology_Biological_Process'
  ))) %>%
  ungroup()
```


### B Cell

```{r b-unique}
unique_susie_fits$fit[[1]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### CD4T Cell

```{r cd4t-unique}
unique_susie_fits$fit[[2]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### CD8T Cell

```{r cd8t-unique}
unique_susie_fits$fit[[3]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### Mono

```{r mono-unique}
unique_susie_fits$fit[[4]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```

### NK

```{r nk-unique}
#| eval: True 
unique_susie_fits$fit[[5]] %>%
  make_cs_tbl_nested() %>%
  nested_reactable()
```






